Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-08-07 submission/manuscript.Rmd
P3 Ada discussion edits

**The number OTUs positively associated with SCFA concentration were similar between individuals without tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without tumors and patients with adenomas or carcinomas in SCFA concentrations or genes encoding enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations. The main goal of this analysis was to identify if there were OTUs that were significantly associated with SCFA concentrations and if this was different between individuals that did not have tumors, had an adenoma, or had a carcinoma. To accomplish this we used Spearmans rho, a non-parametric measure of association, and tested if there was a correlation that was signficantly greater than zero. We found that taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* dominated statistically significant OTU correlations [Figure 3 & Table S5]. There was a noticeably higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3]. In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas [Figure 3]. Patients with adenomas also had more positive correlations between OTUs and SCFA concentrations, but their total number was more similar to individuals without tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3]. Additionally, the number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3]. Finally, when we used high/low SCFA groups based on the overall median concentration for each SCFA instead of SCFA concentrations a similar pattern was still observed [Figure S1 & Table S6]. Overall, these results suggest that the resident taxa that may change the most due to colon tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

**SCFA concentrations do not replace important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** Despite the lack of difference in positive correlations between OTUs and SCFA concentrations between individuals with and without tumors, OTUs associated with SCFA concentrations could still be the most important variables to Random Forest models built to classify patients with adenomas or carcinomas. We tested this by using the Random Forest algorithm to build models with OTU abundance data or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples. With these models we compared whether OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained in the OTU and SCFA model. Additionally, we also compared whether any of the important OTUs that remained also had significant correlations with SCFA concentrations. Both our adenoma and carcinoma models classified patients with a similar degree of success, as measured by the area under the curve (AUC) (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4A & 4D]. After the addition of SCFA concentrations to the adenoma or carcinoma models, many OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained as important variables to the model [Figure Figure 4B-C & 4E-F]. After adding SCFA concentrations to the adenoma model, there were only `r length(unique(sig_assoc_top10_model %>% filter(model == "adn_full") %>% pull(otu)))` OTUs significantly associated with SCFA concentration that were part of the top 10 most important variables and both were postively associated with acetate, butyrate, or propionate concentrations. When SCFA concentrations were added to the carcinoma model, only `r length(unique(sig_assoc_top10_model %>% filter(model == "crc_full") %>% pull(otu)))` OTU that was associated with SCFA concentration remained as part of the top 10 most important variables and it was negatively associated with acetate and butyrate concentrations. In combination with the previous results on OTU correlations, these observations provide additional evidence that the resident taxa that are associated with protection against tumorigeneis are not ones associated with acetate, butyrate, or propionate production.

@@

In addition to not having a response to fiber supplementation due to lack of the required microbes, it is also possible that there are specific instances of colorectal cancer where SCFAs may be beneficial. One limitation of current research into the effect of SCFAs in CRC has been that all tumors are treated as the same type. However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals. Similar to the idea of specific immunotherapy for specific tumors [@Thomas2016], SCFAs may have beneficial effects for specific types of colorectal tumors. Future research will need to test if this is a valid hypothesis. Regardless of this limitation, our results in combination to previous randomized controlled trials on fiber supplementation suggests that using SCFAs as a general treatment for colorectal cancer is unlikely to provide a reduction in tumorigenesis.

>>>>

**The number of OTUs positively associated with SCFA concentration were similar between individuals without tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without tumors and patients with adenomas or carcinomas in SCFA concentrations or genes encoding enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations. The main goal of this analysis was to identify if there were OTUs that were significantly associated with SCFA concentrations and if this was different between individuals that did not have tumors, had an adenoma, or had a carcinoma. To accomplish this we used Spearmans rho, a non-parametric measure of association, and tested if there was a correlation that was signficantly greater than zero. We found that taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* dominated statistically significant OTU correlations [Figure 3 & Table S5]. There was a noticeably higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3]. In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas [Figure 3]. Patients with adenomas also had more positive correlations between OTUs and SCFA concentrations, but their total number was more similar to individuals without tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3]. Additionally, the number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3]. Finally, when we used high/low SCFA groups based on the overall median concentration for each SCFA instead of SCFA concentrations a similar pattern was still observed [Figure S1 & Table S6]. Overall, these results suggest that the resident taxa that may change the most due to colon tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

**SCFA concentrations do not replace important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** Despite the lack of difference in positive correlations between OTUs and SCFA concentrations between individuals with and without tumors, OTUs associated with SCFA concentrations could still be the most important variables to Random Forest models built to classify patients with adenomas or carcinomas. We tested this by using the Random Forest algorithm to build models with OTU abundance data or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples. With these models we compared whether OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained when SCFA concentrations were added to the model. Additionally, we also compared whether any of the important OTUs that remained also had significant correlations with SCFA concentrations. Both our adenoma and carcinoma models classified patients with a similar degree of success, as measured by the area under the curve (AUC) (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4A & 4D]. After the addition of SCFA concentrations to the adenoma or carcinoma models, many OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained as important variables to the model [Figure Figure 4B-C & 4E-F]. After adding SCFA concentrations to the adenoma model, there were only `r length(unique(sig_assoc_top10_model %>% filter(model == "adn_full") %>% pull(otu)))` OTUs significantly associated with SCFA concentration that were part of the top 10 most important variables and both were postively associated with acetate, butyrate, or propionate concentrations. When SCFA concentrations were added to the carcinoma model, only `r length(unique(sig_assoc_top10_model %>% filter(model == "crc_full") %>% pull(otu)))` OTU that was associated with SCFA concentration remained as part of the top 10 most important variables and it was negatively associated with acetate and butyrate concentrations. In combination with the previous results on OTU correlations, these observations provide additional evidence that the resident taxa that are associated with protection against tumorigeneis are not ones associated with acetate, butyrate, or propionate production.

@@

Another possible alternative explanation as to why no difference in SCFA concentration between individuals without tumors and patients with adenomas or carcinomas was observed, could be because only certain types of colorectal cancers are affected by SCFAs. One limitation of current research into the effect of SCFAs and the microbiota in CRC has been that all tumors are treated as the same type. However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in certain subsets of individuals. Similar to the idea of using immunotherapy as a targeted treatment option for specific tumors [@Thomas2016], SCFAs may have beneficial effects for distinct types of colorectal tumors. Future research will need to test if this is a valid hypothesis. Regardless of this limitation, our results in combination to previous randomized controlled trials on fiber supplementation [@Schatzkin2000; @Yao2017] suggests that using SCFAs as a general treatment for colorectal cancer is unlikely to provide a reduction in tumorigenesis.

### 2018-08-07 submission/manuscript.tex
P3 Ada discussion edits

\textbf{The number OTUs positively associated with SCFA concentration
were similar between individuals without tumors and patients with
adenomas or carcinomas.} Having found no difference between individuals
without tumors and patients with adenomas or carcinomas in SCFA
concentrations or genes encoding enzymes involved with SCFA synthesis,
we next investigated if specific OTUs correlated with SCFA
concentrations. The main goal of this analysis was to identify if there
were OTUs that were significantly associated with SCFA concentrations
and if this was different between individuals that did not have tumors,
had an adenoma, or had a carcinoma. To accomplish this we used Spearmans
rho, a non-parametric measure of association, and tested if there was a
correlation that was signficantly greater than zero. We found that taxa
from \emph{Clostridiales}, \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} dominated statistically significant OTU
correlations {[}Figure 3 \& Table S5{]}. There was a noticeably higher
number of significant negative correlations associated with patients
with adenomas for all SCFAs tested {[}Figure 3{]}. In particular, OTUs
from the \emph{Ruminococcaceae} family had the largest share of these
negative correlations within patients with adenomas {[}Figure 3{]}.
Patients with adenomas also had more positive correlations between OTUs
and SCFA concentrations, but their total number was more similar to
individuals without tumors or patients with carcinomas versus the
analogous comparison for the number of negative correlations {[}Figure
3{]}. Additionally, the number of positive correlations between OTUs and
SCFA concentrations was similar between individuals without tumors and
patients with a carcinoma {[}Figure 3{]}. Finally, when we used high/low
SCFA groups based on the overall median concentration for each SCFA
instead of SCFA concentrations a similar pattern was still observed
{[}Figure S1 \& Table S6{]}. Overall, these results suggest that the
resident taxa that may change the most due to colon tumors may not be
ones that are responsible for the production of acetate, butyrate, or
propionate.

\textbf{SCFA concentrations do not replace important
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
OTUs in Random Forest models built to classify tumors.} Despite the lack
of difference in positive correlations between OTUs and SCFA
concentrations between individuals with and without tumors, OTUs
associated with SCFA concentrations could still be the most important
variables to Random Forest models built to classify patients with
adenomas or carcinomas. We tested this by using the Random Forest
algorithm to build models with OTU abundance data or OTU abundances and
SCFA concentrations to classify normal versus adenoma and normal versus
carcinoma fecal samples. With these models we compared whether OTUs with
taxonomic classification to \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} remained in the OTU
and SCFA model. Additionally, we also compared whether any of the
important OTUs that remained also had significant correlations with SCFA
concentrations. Both our adenoma and carcinoma models classified
patients with a similar degree of success, as measured by the area under
the curve (AUC) (P-value \textgreater{} 0.05) {[}Figure 4A \& 4D{]}.
After the addition of SCFA concentrations to the adenoma or carcinoma
models, many OTUs with taxonomic classification to \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} remained as important
variables to the model {[}Figure Figure 4B-C \& 4E-F{]}. After adding
SCFA concentrations to the adenoma model, there were only 2 OTUs
significantly associated with SCFA concentration that were part of the
top 10 most important variables and both were postively associated with
acetate, butyrate, or propionate concentrations. When SCFA
concentrations were added to the carcinoma model, only 1 OTU that was
associated with SCFA concentration remained as part of the top 10 most
important variables and it was negatively associated with acetate and
butyrate concentrations. In combination with the previous results on OTU
correlations, these observations provide additional evidence that the
resident taxa that are associated with protection against tumorigeneis
are not ones associated with acetate, butyrate, or propionate
production.

@@

In addition to not having a response to fiber supplementation due to
lack of the required microbes, it is also possible that there are
specific instances of colorectal cancer where SCFAs may be beneficial.
One limitation of current research into the effect of SCFAs in CRC has
been that all tumors are treated as the same type. However, there are
known differences in the types of mutations that occur (25) and treating
all tumors as equal may actually hide any benefit that could be found in
specific subsets of individuals. Similar to the idea of specific
immunotherapy for specific tumors (26), SCFAs may have beneficial
effects for specific types of colorectal tumors. Future research will
need to test if this is a valid hypothesis. Regardless of this
limitation, our results in combination to previous randomized controlled
trials on fiber supplementation suggests that using SCFAs as a general
treatment for colorectal cancer is unlikely to provide a reduction in
tumorigenesis.

>>>>

\textbf{The number of OTUs positively associated with SCFA concentration
were similar between individuals without tumors and patients with
adenomas or carcinomas.} Having found no difference between individuals
without tumors and patients with adenomas or carcinomas in SCFA
concentrations or genes encoding enzymes involved with SCFA synthesis,
we next investigated if specific OTUs correlated with SCFA
concentrations. The main goal of this analysis was to identify if there
were OTUs that were significantly associated with SCFA concentrations
and if this was different between individuals that did not have tumors,
had an adenoma, or had a carcinoma. To accomplish this we used Spearmans
rho, a non-parametric measure of association, and tested if there was a
correlation that was signficantly greater than zero. We found that taxa
from \emph{Clostridiales}, \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} dominated statistically significant OTU
correlations {[}Figure 3 \& Table S5{]}. There was a noticeably higher
number of significant negative correlations associated with patients
with adenomas for all SCFAs tested {[}Figure 3{]}. In particular, OTUs
from the \emph{Ruminococcaceae} family had the largest share of these
negative correlations within patients with adenomas {[}Figure 3{]}.
Patients with adenomas also had more positive correlations between OTUs
and SCFA concentrations, but their total number was more similar to
individuals without tumors or patients with carcinomas versus the
analogous comparison for the number of negative correlations {[}Figure
3{]}. Additionally, the number of positive correlations between OTUs and
SCFA concentrations was similar between individuals without tumors and
patients with a carcinoma {[}Figure 3{]}. Finally, when we used high/low
SCFA groups based on the overall median concentration for each SCFA
instead of SCFA concentrations a similar pattern was still observed
{[}Figure S1 \& Table S6{]}. Overall, these results suggest that the
resident taxa that may change the most due to colon tumors may not be
ones that are responsible for the production of acetate, butyrate, or
propionate.

\textbf{SCFA concentrations do not replace important
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
OTUs in Random Forest models built to classify tumors.} Despite the lack
of difference in positive correlations between OTUs and SCFA
concentrations between individuals with and without tumors, OTUs
associated with SCFA concentrations could still be the most important
variables to Random Forest models built to classify patients with
adenomas or carcinomas. We tested this by using the Random Forest
algorithm to build models with OTU abundance data or OTU abundances and
SCFA concentrations to classify normal versus adenoma and normal versus
carcinoma fecal samples. With these models we compared whether OTUs with
taxonomic classification to \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} remained when SCFA
concentrations were added to the model. Additionally, we also compared
whether any of the important OTUs that remained also had significant
correlations with SCFA concentrations. Both our adenoma and carcinoma
models classified patients with a similar degree of success, as measured
by the area under the curve (AUC) (P-value \textgreater{} 0.05)
{[}Figure 4A \& 4D{]}. After the addition of SCFA concentrations to the
adenoma or carcinoma models, many OTUs with taxonomic classification to
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
remained as important variables to the model {[}Figure Figure 4B-C \&
4E-F{]}. After adding SCFA concentrations to the adenoma model, there
were only 2 OTUs significantly associated with SCFA concentration that
were part of the top 10 most important variables and both were postively
associated with acetate, butyrate, or propionate concentrations. When
SCFA concentrations were added to the carcinoma model, only 1 OTU that
was associated with SCFA concentration remained as part of the top 10
most important variables and it was negatively associated with acetate
and butyrate concentrations. In combination with the previous results on
OTU correlations, these observations provide additional evidence that
the resident taxa that are associated with protection against
tumorigeneis are not ones associated with acetate, butyrate, or
propionate production.

@@

Another possible alternative explanation as to why no difference in SCFA
concentration between individuals without tumors and patients with
adenomas or carcinomas was observed, could be because only certain types
of colorectal cancers are affected by SCFAs. One limitation of current
research into the effect of SCFAs and the microbiota in CRC has been
that all tumors are treated as the same type. However, there are known
differences in the types of mutations that occur (25) and treating all
tumors as equal may actually hide any benefit that could be found in
certain subsets of individuals. Similar to the idea of using
immunotherapy as a targeted treatment option for specific tumors (26),
SCFAs may have beneficial effects for distinct types of colorectal
tumors. Future research will need to test if this is a valid hypothesis.
Regardless of this limitation, our results in combination to previous
randomized controlled trials on fiber supplementation (20, 23) suggests
that using SCFAs as a general treatment for colorectal cancer is
unlikely to provide a reduction in tumorigenesis.
